---
title: "GAM power analysis"
author: "Chris Terry"
output: 
  html_document: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This document conducts a basic power analysis through simulation of the capacity for GAMs to identify higher-order interactions in  response surface of plausible *Drosophila* experimental data. 

The core objective is to estimate the finite rate of population increase, **\(\lambda\)**, given the focal population density $N_{i,t}$, the competitor density $N_{j,t}$ and the temperature $T_t$. 


$$
N_{i,t+1}=N_{i,t}\lambda_{i,t}
$$
$$
\lambda_{i,t}=f_i(N_{i,t}, N_{j,t}, T_t)
$$
The core result is that  that several thousand observations are likely to be sufficient to be able to detect 3-way interactions that cause +/- 20% impact, given plausible levels of experimental noise. 


# Generating simulated data. 

## Model
Data generating models are derived from the results of Terry (2025, https://doi.org/10.1111/ele.70047), in which extensive model selection was used to fit models of thermal performance and competition for two fly species *Drosophila pandora* and *Drosophila pallidifrons*. The generations are discrete - adults are able to lay eggs for a fixed period then removed. After 10 days development, the subsequent emergent adults are then counted. 

The best fit model used described the error with a zero-inflated negative binomial model. 

In that experiment Drosophila vials were used, whereas in this trial I will instead use larger Drosophila bottles, that will be able to support a larger population, however for the purposes of power analysis the underlying dynamics is likely to be similar. 




The best model (selected via ELPD-LOO) for the expected values was:

$$\lambda =\underbrace{B_0T(T-T_{\min})\sqrt{T_{\max}-T}}_{\text{thermal performance}}
\times \underbrace{\frac{1}{1+\alpha_{ii}N_i+\alpha_{ij}(T)N_j}}_{\text{Beverton-Holt Competition}}
$$


where the competition was linearly dependent on the temperature (scaled to aid fitting)

$$\alpha_{ij}(T)=\alpha_{ij0}+\alpha_{ijT}(T-20).$$


and draws are from a zero-inflated negative-binomial model. 



```{R message=FALSE, warning=FALSE}
library(tidyverse)
library(mgcv)   
library(gratia) # for visualising GAMs
```

## Parameters 

Parameters for one of the species, *D. pandora*, are taken from the posterior mean values from the archived model fits (https://github.com/jcdterry/MCT_Fly_Val_public/blob/main/Analysis/4_Inferences-from-best-fit-models.html) .

```{R}
# Parameters from the PAN best-fit model
B0        <- 0.01
Tmin      <- 17.82
Tmax      <- 33.59

alphaii   <- 0.17
alphaij0  <- 0.74
alphaijT  <- -0.06

# Negative-binomial / zero-inflation parameters
shape_intercept <- -1.14
shape_focal     <- 0.05
zi              <- 0.04
```

## Potential Experimental Design

The choice of levels will be adaptive, with later blocks emphasising areas of greater change to maximise the information from the sampling. To represent that in this analysis, the sampling at lower levels is increased. 

```{R}
# Experimental design
density_values<- c(0,0,  # extra replication at zero
                   seq(0, 20, by= 4), # dense low-density coverage
                   seq(30, 100, by =10)) # # broader high-density coverage

dat <- expand_grid(temp = seq(22, 28, length.out = 7),
                      focal_density = density_values,
                      comp_density =density_values,
                      replicate = 1:6)
```

## Underlying biological response surface

```{R}
# temp-dependent intrinsic growth
therm_perf <-
  B0 * dat$temp * (dat$temp - Tmin) * sqrt(pmax(Tmax - dat$temp, 0))

# temp-dependent interspecific competition
alphaij <-alphaij0 + alphaijT * (dat$temp - 20)

## Expected value at next generation
mu <-dat$focal_density *therm_perf/( 1+alphaii*dat$focal_density+alphaij*dat$comp_density)
```

## Adding noise to the observations

Three points complicate generating an appropriate amount of noise on top of the expected values: 

- In the planned experiments, I will use Drosophila bottles, rather than vials, which are somewhat larger. Increasing the size of the container will somewhat reduce the demographic noise as the carrying capacity can be mostly controlled by the amount of food available. Based on previous work, populations may be approximately 4x larger than in the vials. I have accounted for this by scaling the size parameter with this. 

- In the 2025 paper, the best fit model identified that dispersion scaled with the population density. This is included by linking the shape parameter to the focal density. 

- A  feature of Drosophila assays in the vials is the sizeable fraction of zero's, where the line collapses (this can be due to too high a density, delaying development, or something causing no eggs to be laid). In bottles this is likely to be less pronounced than in vials, but here maintaining the  zero-inflation observed in the experiment (4%) as a conservative test.


```{R}
shape_vials <- exp( shape_intercept + shape_focal * dat$focal_density)
shape_bottle <- 4 * shape_vials # shift in size of bottle. 

# Structural zero component
structural_zero <- runif(nrow(dat)) < zi

# Generate next-generation population size
dat$focal_next_generation <- 0
positive <- (!structural_zero) & (mu > 0)

set.seed(1)
dat$focal_next_generation[positive] <-rnbinom( sum(positive),
                                               size = shape_bottle[positive],
                                               mu   = mu[positive])
```

## Visualisation of noise

```{R}
dat %>%
  filter( temp == 25, comp_density == 20) %>%
  ggplot( aes( x = focal_density, y = focal_next_generation))+
  geom_point()+
  theme_classic()+
  ggtitle( label = 'Slice through observation grid,\nwhere T = 25, Nj = 20')

```


## Plotting Response Surfaces

Here plotting out expected value surfaces. 

```{R}
GridPoints<- c(0, 2,10, 20, 40, 60, 80, 100)
breaks_loglambda <- seq(-4, 4, by = 0.5)

expand_grid(temp = c(22, 25, 28),
            focal_density = seq(1, 100, length.out = 100),
            comp_density = seq(0, 100, length.out = 100)) %>%
  mutate( therm_perf = B0 * temp * (temp - Tmin) *sqrt(pmax(Tmax - temp, 0)),
          alphaij =alphaij0 + alphaijT * (temp - 20),
          lambda = therm_perf / (1 + alphaii * focal_density + alphaij * comp_density)) %>%
  mutate( focal_density_plot = sqrt(focal_density),
          comp_density_plot = sqrt(comp_density),
          log_lambda = log(lambda))%>%
  ggplot(  aes(  x = comp_density_plot,
                 y = focal_density_plot,
                 z = log_lambda )) +
  geom_contour_filled( breaks = breaks_loglambda) +
  scale_fill_brewer(palette = "RdBu",
                    direction = -1,
                    name = expression("Per-capita growth  log(" * lambda * ")"))+
  facet_wrap(~ temp, nrow = 1,
             labeller = label_bquote( .(temp) ~ degree*C)) +
  scale_x_continuous( breaks = sqrt(GridPoints),
                      labels = GridPoints ) +
  scale_y_continuous( breaks = sqrt(GridPoints),
                      labels = c(GridPoints)) +
  labs(x = "Competitor density",
       y = "Focal density",
       fill = "log(λ)") +
  theme_classic()

```

# Fitting GAMs with differnt orders of interactions.

Here fitting 3 demonstration models of $N_{i,t+1}$ that include just direct effects of the three predictors (m1), pairwise interactions (m2), and a three-way interaction (m3). 

The model includes a log-offset and a negatove-binomial distribution so the model can be directly fit to the observation data, and fit the per-capita population growth surface. 

i.e.: 

$$
\log(E[N_{t+1}])
=
\underbrace{\log(N_t)}_{\text{offset}}
+
\underbrace{f(T,N_t,N_j)}_{\text{GAM learns this}}.
$$

and hence: 

$$
\lambda =
\frac{E[N_{t+1}]}{N_t}
=
\exp[f(T,N_t,N_j)].
$$

Note that the underlying model that the data was generated from does not include a direct three-way interaction. So if the model is behaving well, `m3` should not explain more of the variance than `m2` 

```{R}
dat_nb <- filter(dat, focal_density > 0)### Don't fit to clear cases of zero. 

m1 <- gam(focal_next_generation ~
            offset(log(focal_density)) +
            s(temp, k = 7) +
            s(focal_density, k = 8) +
            s(comp_density, k = 8),
          family = nb(), method = "REML", data = dat_nb)

m2 <- gam(  focal_next_generation ~
              offset(log(focal_density)) +
              
              s(temp, k = 7) +
              s(focal_density, k = 8) +
              s(comp_density, k = 8) +
              ## two way interactions:
              ti(temp, focal_density, k = c(6, 6)) +
              ti(temp, comp_density, k = c(6, 6)) +
              ti(focal_density, comp_density, k = c(6, 6)),
            
            family = nb(), method = "REML", data = dat_nb)

m3 <- gam( focal_next_generation ~
             offset(log(focal_density)) +
             s(temp, k = 7) +
             s(focal_density, k = 8) +
             s(comp_density, k = 8) +
             ti(temp, focal_density, k = c(6, 6)) +
             ti(temp, comp_density, k = c(6, 6)) +
             ti(focal_density, comp_density, k = c(6, 6)) +
             ## three-way interactions
             ti(temp, focal_density, comp_density, k = c(5, 5, 5)),
           
           family = nb(), method = "REML", data = dat_nb)
```


```{r}
summary(m1)
summary(m2)
summary(m3)

par(mfrow= c(2,2))
gam.check(m3)

```

### Slice Plots to inspect


```{R}
slice_dat <- expand_grid(temp = c(22, 25, 28),
                         comp_density = c(0, 20, 60),
                         focal_density = seq(1, 100, length.out = 200)) %>%
  mutate(therm_perf = B0 * temp * (temp - Tmin) * sqrt(pmax(Tmax - temp, 0)),
         alphaij =alphaij0 + alphaijT * (temp - 20),
         lambda_true = therm_perf / (1 +  alphaii * focal_density +  alphaij * comp_density),
         log_lambda_true = log(lambda_true))

slice_dat$N_next_hat <- predict(m3,newdata = slice_dat,type = "response")
slice_dat$lambda_hat <- slice_dat$N_next_hat / slice_dat$focal_density
slice_dat$log_lambda_hat <-log(slice_dat$lambda_hat)


slice_dat$N_next_hat <- predict(m2, newdata = slice_dat, type = "response")

slice_dat <- slice_dat %>%
  mutate(lambda_hat = N_next_hat / focal_density,
    log_lambda_hat = log(lambda_hat))

ggplot(slice_dat, aes(x = focal_density)) +
  geom_line(aes(y = log_lambda_true),linewidth = 1) +
  geom_line( aes(y = log_lambda_hat), linetype = 2, linewidth = 0.8, col = 'red') +
  facet_grid(temp ~ comp_density,
    labeller = labeller( temp = function(x) paste0(x, " °C"),
      comp_density =function(x) paste0("Competitor = ", x))) +
  geom_hline(yintercept = 0,linewidth = 0.3) +
  labs(x = "Focal density",y = expression(log(lambda)) ) +
  theme_classic()

```

# Testing effect of increasing sample size. 

Here varying the number of replicates per level combination, regenerating independent datasets, and then refitting the 'm2' (pairwise interactions) 

## Function:

```{R}

fit_one <- function(n_rep, container = c('bottle', 'vial')[1]) {
  
  ### Generating data
  dat <- expand_grid(temp = seq(22, 28, length.out = 7),
                     focal_density = density_values,
                     comp_density =density_values,
                     replicate = 1:n_rep) %>% # number of replicates per level
    filter(focal_density != 0 )
  
  ### Underlying biological response surface
  # temp-dependent intrinsic growth
  therm_perf <-B0 * dat$temp * (dat$temp - Tmin) * sqrt(pmax(Tmax - dat$temp, 0))
  alphaij <-alphaij0 + alphaijT * (dat$temp - 20)
  mu <-dat$focal_density *therm_perf/( 1+alphaii*dat$focal_density+alphaij*dat$comp_density)
  # Noise
  shape_bottle <- 4 * exp( shape_intercept + shape_focal * dat$focal_density)
  shape_vial <- 4 * exp( shape_intercept + shape_focal * dat$focal_density)
  structural_zero <- runif(nrow(dat)) < zi  # Structural zero component
  # Generate next-generation population size
  dat$focal_next_generation <- 0
  positive <- (!structural_zero) & (mu > 0)
  
  if( container == 'bottle'){
    dat$focal_next_generation[positive] <- rnbinom( sum(positive),
                                                    size = shape_bottle[positive],
                                                    mu   = mu[positive])
  }
  if( container == 'vial'){
    dat$focal_next_generation[positive] <- rnbinom( sum(positive),
                                                    size = shape_vial[positive],
                                                    mu   = mu[positive]) 
  }
  
  
  # Pairwise-context GAM
  mod <- gam(focal_next_generation ~
               offset(log(focal_density)) +
               s(temp, k = 7) +
               s(focal_density, k = 8) +
               s(comp_density, k = 8) +
               ti(temp, focal_density, k = c(6, 6)) +
               ti(temp, comp_density,  k = c(6, 6)) +
               ti(focal_density, comp_density, k = c(6, 6)),
             family = nb(),
             method = "REML",data = dat)
  
  # Extract smooth-term table
  sm <- summary(mod)$s.table
  
  tibble(n_rep = n_rep,
         total_n = nrow(dat),
         p_temp_comp = sm["ti(temp,comp_density)", "p-value"],
         p_temp_focal =  sm["ti(temp,focal_density)", "p-value"],
         p_focal_comp = sm["ti(focal_density,comp_density)", "p-value"],
         detected =p_temp_comp < 0.05,
         deviance_explained =summary(mod)$dev.expl
  )
}


fit_one_safe <- safely(fit_one)
```

## Fitting 10 times:

It appears that  even 1 replicate per level combination (n =1456) can reliably detect the 2-way interaction . 

```{R}
set.seed(123)
n_reps <- c(1)
n_sims <- 10

results <- map_dfr(rep(n_reps,each =n_sims),
                   fit_one_safe )
```

# Testing recovery of a three-way interaction

What if a there was a three-way interaction between temperature and two species density? How many observations would be needed to reliably pull this out? This wasn't directly tested for in the original modelling, but can be added into the simulator.   

The three way interaction here is: "The per-capita competitive effect of species \(j\) can differ by up to ±20% according to the joint temperature and focal-density context." For this reason, it is fit using scaled temperature and focal density. 


```{r}
# Values to centre contextual variables
T_centre <- 25
T_scale <- 3
N_centre <- mean(density_values)
N_scale <- max(abs(density_values - N_centre))

fit_model_3way <- function(n_rep, beta3, container = c('bottle', 'vial')[1]) {
  ## Output progress
  print(n_rep )
  print(Sys.time())
  
  ### Generating data
  dat <- expand_grid(temp = seq(22, 28, length.out = 7),
                     focal_density = density_values,
                     comp_density =density_values,
                     replicate = 1:n_rep) %>% # number of replicates per level
    filter(focal_density != 0 )
  
  ### Underlying biological response surface
  # temp-dependent intrinsic growth
  therm_perf <-B0 * dat$temp * (dat$temp - Tmin) *sqrt(pmax(Tmax - dat$temp, 0))
  # Existing temperature-dependent competition
  alphaij_base <- alphaij0 +  alphaijT * (dat$temp - 20)
  
   # Standardised contextual variables
  T_scaled <-  (dat$temp - T_centre) / T_scale
  N_scaled <- (dat$focal_density - N_centre) / N_scale
  
  #####################################
  # Three-way modification
  threeway_multiplier <- 1 + beta3 * T_scaled * N_scaled
  alphaij <- alphaij_base * threeway_multiplier
  ####################################
  
  # Population growth
  mu <- dat$focal_density *  therm_perf /(1 +alphaii * dat$focal_density + alphaij * dat$comp_density)
  
  # Noise
  shape_bottle <- 4 * exp( shape_intercept + shape_focal * dat$focal_density)
  shape_vial <- 4 * exp( shape_intercept + shape_focal * dat$focal_density)
  structural_zero <- runif(nrow(dat)) < zi  # Structural zero component
  # Generate next-generation population size
  dat$focal_next_generation <- 0
  positive <- (!structural_zero) & (mu > 0)
  
  if( container == 'bottle'){
    dat$focal_next_generation[positive] <- rnbinom( sum(positive),
                                                    size = shape_bottle[positive],
                                                    mu   = mu[positive])
  }
  if( container == 'vial'){
    dat$focal_next_generation[positive] <- rnbinom( sum(positive),
                                                    size = shape_vial[positive],
                                                    mu   = mu[positive]) 
  }
  

  # Pairwise-context GAM
  mod <- gam(focal_next_generation ~
               offset(log(focal_density)) +
               s(temp, k = 7) +
               s(focal_density, k = 8) +
               s(comp_density, k = 8) +
               ti(temp, focal_density, k = c(6, 6)) +
               ti(temp, comp_density,  k = c(6, 6)) +
               ti(focal_density, comp_density, k = c(6, 6)) +
               ti(temp, focal_density, comp_density,
                  k = c(5, 5, 5)),
             family = nb(),
             method = "REML",data = dat)
  
  # Extract smooth-term table
  sm <- summary(mod)$s.table
  
  tibble(n_rep = n_rep,
         total_n = nrow(dat),
         beta3= beta3, 
         p_temp_comp = sm["ti(temp,comp_density)", "p-value"],
         p_temp_focal =  sm["ti(temp,focal_density)", "p-value"],
         p_focal_comp = sm["ti(focal_density,comp_density)", "p-value"],
         p_3way = sm["ti(temp,focal_density,comp_density)", "p-value"],
         detected =p_3way < 0.05,
         deviance_explained =summary(mod)$dev.expl
  )
}


```

## Runnning detection of 3-way interactions

```{r eval = FALSE}

### Null Test Without HOIs
null_test <- map_dfr(1:30,~ fit_model_3way( n_rep = 4,beta3 = 0))

write_csv(null_test, 'null_test.csv' )

n_reps <- c(1,2,4,8)
n_sims <- 20


### Test *With* HOIs at 20% effect
results_beta3_0.2 <- map_dfr(rep(n_reps,each =n_sims),
                             fit_model_3way , beta3 = 0.2)

write_csv(results_beta3_0.2, 'results_beta3_0.2.csv' )
```

```{r}
null_test<- read_csv('null_test.csv' )
results_beta3_0.2<- read_csv('results_beta3_0.2.csv' )

mean(null_test$detected)
null_test$p_3way %>% hist(breaks = 20)


results_beta3_0.2 %>%
  ggplot(aes( x=total_n , y =p_3way  ))+
  geom_point()+
  geom_hline( yintercept = 0.05)+
  theme_classic()

results_beta3_0.2 %>%
  count( total_n, detected)

```


# Repeating without vial-to-bottle scaling

## Visualisation of noise

```{R}
dat_vial <- dat

# Structural zero component
structural_zero <- runif(nrow(dat_vial)) < zi

# Generate next-generation population size
dat_vial$focal_next_generation <- 0
positive <- (!structural_zero) & (mu > 0)

set.seed(1)
dat_vial$focal_next_generation[positive] <-rnbinom( sum(positive),
                                               size = shape_vials[positive],
                                               mu   = mu[positive])

dat_vial %>%
  filter( temp == 25, comp_density == 20) %>%
  ggplot( aes( x = focal_density, y = focal_next_generation))+
  geom_point()+
  theme_classic()+
  ggtitle( label = 'Slice through observation grid,\nwhere T = 25, Nj = 20\n(Vial-Scale Noise)')
```

# Testing effect of increasing sample size on recovery of a three-way interaction

## Runnning detection of 3-way interactions

```{r eval = FALSE}
null_test <- map_dfr(1:30,~ fit_model_3way( n_rep = 4,beta3 = 0, container = 'vial'))

write_csv(null_test, 'null_test_noScaling.csv' )

n_reps <- c(1,2,4,8)
n_sims <- 20

results_beta3_0.2 <- map_dfr(rep(n_reps,each =n_sims),
                             fit_model_3way , beta3 = 0.2, container = 'vial')

write_csv(results_beta3_0.2, 'results_beta3_0.2_noScaling.csv' )
```

With 'Vial-Scale' noise, false-positive crop up at a similiar rate (4/30). Successful detection of the moderate-strength three-way interaction is a little weaker, with the break even point coming between 3000 and 5000 contributing data points. 

```{r}
null_test<- read_csv('null_test_noScaling.csv' )
results_beta3_0.2<- read_csv('results_beta3_0.2_noScaling.csv' )

mean(null_test$detected)
null_test$p_3way %>% hist(breaks = 20)


results_beta3_0.2 %>%
  ggplot(aes( x=total_n , y =p_3way  ))+
  geom_point()+
  geom_hline( yintercept = 0.05)+
  theme_classic()

results_beta3_0.2 %>%
  count( total_n, detected)
```

# Session Info

```{r}
sessionInfo()
```
